Loop-Aware Optimizations in PyPy's Tracing JIT

### 2011-06-21 talk/iwtc11/paper.tex
fix two XXXs, start adding some minimal notes to the related work section

XXX "bridge" is not a standard term

XXX loop peeling does a lot more than loop-invariant code motion
% take this loop as an example:
% [i1, i2]
% i3 = i1 + 1
% i4 = i2 + 1
% escape(i4)
% jump(i2, i3)
% none of the operations is loop-invariant, but loop peeling will still remove the second addition

@@

Loop peeling is achieved by appending an copy of the traced iteration at
the end of itself. See Figure~\ref{fig:overview} for an illustration.
The first part (called \emph{preamble}) finishes with the jump the the second part
(called the \emph{peeled loop}). The second part end with the jump to itself. This way
the preamble will be executed only once while the peeled loop will
be used for every further iteration. New variable names have to be
introduced in the entire copied trace in order to maintian the SSA-property.
When applying the following optimizations to this two-iteration trace
some care has to taken as to how the arguments of the two
\lstinline{jump} operations and the input arguments of the peeled loop are
treated. It has to be ensured that the peeled loop stays a proper
trace in the sense that the operations within it only operates on
variables that are either among its input arguments 
or produced within the peeled loop. To ensure this we need
to introduce a bit of formalism.

@@

XXX should we have a mention where in the previous papers those optimizations
are discussed?

@@

The loop peeling optimization was implemented in the PyPy
framework. That means that the JIT-compilers generated for all
interpreters implemented within PyPy now can take advantage of
it. Benchmarks have been executed for a few different interpreters and
we see improvements in several cases. The ideal loop for this optimization
would be short numerical calculations with no failing guards and no
external calls. Larger loops involving many operations on complex objects
typically benefit less from it. Loop peeling never makes runtime performance worse, in
the worst case the peeled loop is exactly the same as the preamble. Therefore we
chose to present benchmarks of small numeric kernels where loop peeling can show
its use.

@@

>>>>

% loop peeling does a lot more than loop-invariant code motion
% take this loop as an example:
% [i1, i2]
% i3 = i1 + 1
% i4 = i2 + 1
% escape(i4)
% jump(i2, i3)
% none of the operations is loop-invariant, but loop peeling will still remove the second addition

@@

Loop peeling is achieved by appending an copy of the traced iteration at
the end of itself. See Figure~\ref{fig:overview} for an illustration.
The first part (called \emph{preamble}) finishes with the jump the the second part
(called the \emph{peeled loop}). The second part end with the jump to itself. This way
the preamble will be executed only once while the peeled loop will
be used for every further iteration. New variable names have to be
introduced in the entire copied trace in order to maintian the SSA-property.
Note that the peeled loop is not necessary the \emph{first} iteration of the
loop execution, it is general enough to correspond to any iteration of the loop.
However, the peeled loop can then be optimized using the assumption that a
previous iteration has happened.

When applying optimizations to this two-iteration trace
some care has to taken as to how the arguments of the two
\lstinline{jump} operations and the input arguments of the peeled loop are
treated. It has to be ensured that the peeled loop stays a proper
trace in the sense that the operations within it only operates on
variables that are either among its input arguments 
or produced within the peeled loop. To ensure this we need
to introduce a bit of formalism.

@@

@@

The loop peeling optimization was implemented in the PyPy
framework in about 450 lines of RPython code. That means that the JIT-compilers generated for all
interpreters implemented within PyPy now can take advantage of
it. Benchmarks have been executed for a few different interpreters and
we see improvements in several cases. The ideal loop for this optimization
would be short numerical calculations with no failing guards and no
external calls. Larger loops involving many operations on complex objects
typically benefit less from it. Loop peeling never makes runtime performance worse, in
the worst case the peeled loop is exactly the same as the preamble. Therefore we
chose to present benchmarks of small numeric kernels where loop peeling can show
its use.

@@

All the optimizations presented here are completely standard \cite{XXX} [dragon
book or similar].

LuaJIT does the same

SPUR does all these optimizations manually, moves allocations out of loop but
does not explode objects
